---
title: "Filzbach Run"
author: "Russell Dinnage"
date: "Thursday, May 29, 2014"
output: md_document
---

```{r setup, echo=FALSE, cache=TRUE, results='hide', message=FALSE, warning=FALSE}
library(knitr)
source("D:/Users/Dinnage/Projects/PLNT-model/R/PLNT-LLmodel.R")
```

This vignette shows how I fit a Phylogenetic Latent Niche Trait model on a fish dataset that was kindly provided to me by Mattew Helmus.

The first step is to load in the fish dataset. This data consists of the presence or absence of 11 sunfish species measured in 890 lakes through-out "" United States. Additionally there is a phylogenetic tree containing all of these sunfish, and also some environmental data for each lake that we won't make use of initially.

```{r load data, results='asis', cache=TRUE}
## load P/A data
fishPA <- read.csv("D:/Users/Dinnage/Projects/PLNT-model/data/HelmusFishPA.csv")
kable(head(fishPA))
## load phylogenetic tree, which is in the form of a phylogenetic covariance matrix
fishPhylo <- read.csv("D:/Users/Dinnage/Projects/PLNT-model/data/HelmusFishPhyloCV.csv")
rownames(fishPhylo)<-fishPhylo[,1]
fishPhylo<-fishPhylo[,-1]
## visualize phylogenetic covariance matrix (actually a phylogenetic correlation matrix to be exact)
library(lattice)
levelplot(as.matrix(fishPhylo), scales=list(x=list(rot=90)), xlab="", ylab="", col.regions = gray(0:100/100))
```

The fundamental probability of species $i$ existing at a site $j$ is determined by the following equation:

$$ P_{i, j}^{F} = K_{i}\cdot \exp\left (\sum_{d}^{d_{max}} -1\cdot \left [ \frac{E_{d,j} - \phi _{d,i}}{\sigma_{d,i} } \right ]^{2} \right ) $$

This equation simply describes a species probability distribution based on it's hypothetical or latent traits (the $\phi _{d,i}$) and the site's hypothetical or latent environmental optimum (the $E_{d,j}$) for $d$ traits. The function is Gaussian, with a species having a probability of $K_{i}$ when it's traits are exactly on the site's environmental optimum. If the species traits and the site's optimum are difference from each other, the probability declines with distance according to a Gaussian function. $K_{i}$ is contrained to fall between 0 and 1. In this model we are treating species' traits and site's environmental optimums as latent variable, to be estimated by the model, however, if one had data on species' traits, it would be trivial to replace the latent trait parameter vector with a data vector of the measured traits. Likewise, the $E_{d,j}$ could be thought of an environmental property of some kind, and the $\phi _{d,i}$ variables as the species' optimum for those environmental properties, in which case you could substitute measured environmental variables for the fitted parameters $E_{d,j}$. Either way, the model works the same way. Even better would be to use measured environmental variables but also include some latent environmental parameters, which could account for structure in the data unexplained by what was measured.

The above model captures the effect of environmental filters on species, but we also wish to look for the signature of competitive species interactions. To do this, each species' fundamental probability of existing in a site is modified by interactions with other species to create a realized probability of existing. This is achieved by reducing a species' probability proportional to the probability of other species being in that site, weighted by pairwise species competition coefficients. This is what the equation looks like:

$$ P_{i, j}^{R} = \max \left [ \left ( P_{i,j}^{F} - \sum _{k \neq q} \alpha _{k,q}\cdot P_{i,q}^{R} \right ), 0 \right ] $$

This is a highly non-linear equation because it contains the $\max$ function in order to keep all probabilities positive, and also because each species realized probability depends on the realized probability of all other species. There is no simple analytical solution for this type of equation, but it can be solved readily using numerical methods, which is what I do here, using the R package `BB` for numerical solving of non-linear equations.



Once we have calculated the realized probabilities of all species for all sites under a set of parameters, it is simple to calculate the likelihood of that set of parameters. Using the observed data, if a species is present at a site then the likelhood is just $P_{i, j}^{R}$, if a species is not at a site then the likelihood is $1 - P_{i, j}^{R}$. The full likelihood then is simply the product of this likelihood across all species-site combinations.

This is the log likelihood function for the model translated into R code:

```{r ll, eval=FALSE}
ll.PLNT<- function(params, comm, phylo) {
  ## meld phylogenetic correlation matrix with diagonal matrix which represents 
  ## no correlation b/w species to form new hybrid CV matrix
  CVmat<-(params$PhyEff)*phylo + (1-params$PhyEff)*diag(1,params$Ni)
  ## Apply cholesky decomposition of of CV matrix to normally distributed
  ## Phi vector to force values to be (phylogenetically) correlated
  Phi<-llply(params$Phi, function (x) params$Sigi*(x%*%chol(CVmat)))
  E<-llply(params$E, function (x) params$Sigj*x) ## apply environmental variance (Sigj)
  ## Make V vectors into matrices
  Emat<-llply(E,function(x) matrix(rep(x,length.out=params$Nj*params$Ni),
                                          nrow=params$Ni,ncol=params$Nj,byrow=TRUE) )
  ## Make Phi vectors into matrices
  Phimat<-llply(Phi,function(x) matrix(rep(x,length.out=params$Nj*params$Ni),
                                              nrow=params$Ni,ncol=params$Nj))
  ## Make Sig value into matrices
  Sigmat<-rlply(params$Nd,matrix(rep(params$Sig,length.out=params$Nj*params$Ni),
                                 nrow=params$Ni,ncol=params$Nj))
  ## Do matrix function to calculate fundamental probabilities of species in sites
  ## according to model ()
  Pf<-mapply(function(x,y,z) ((x-y)/z)^2, Emat, Phimat, Sigmat, SIMPLIFY=FALSE)
  Pf<-exp((-1)*do.call("+",Pf)) #Pf = Fundamental probability of existence of 
  ## species j (rows) in site q (columns)
  ## Calculate alpha competition coefficients for each competition dimension
  Alphmat<-llply(params$Alph,function(x) ((matrix(rep(x,length.out=params$Ni^2),
                                                  nrow=params$Ni,ncol=params$Ni,byrow=TRUE)-
                   matrix(rep(x,length.out=params$Ni^2),nrow=params$Ni,ncol=params$Ni))/params$w)^2)
  Alphmat<-exp((-1)*do.call("+",Alphmat)) ## multiply competition dimensions
  diag(Alphmat)<-NA
  test<-BBsolve(runif(Ni*Nj,0,1),BBfun,Pf=Pf,Alph=Alphmat,Ni=params$Ni,Nj=params$Nj,
                control=list(trace=TRUE))
  Pr<-matrix(test$par,nrow=params$Ni,ncol=params$Nj) ## Pr = Realized probability of existence 
  ## of species i (rows) in site j (columns)
  LL<-matrix(NA,nrow=params$Ni,ncol=params$Nj)
  LL[comm==1]<-log(Pr) ## Log Likelihood of observing species
  LL[comm==0]<-log(1-Pr) ## Log Likelihood of not observing species
  return(sum(LL)) ## Return full likelihood!
}
## Numerical solver function for probability of a species existing at a site taking into
## account other species and their competitive effects
BBfun<- function(x, Pf, Alph, Ni, Nj){
  xmat<-matrix(x,nrow=Ni,ncol=Nj)
  xnew<-alply(xmat,2,identity)
  alph1<-mapply(function(x,y) x*y, xnew, list(Alph),SIMPLIFY=FALSE)
  alpha<-t(laply(alph1,function(x) aaply(x,1,function(y) sum(y,na.rm=TRUE))))
  y<-pmax(matrix(0,nrow=Ni,ncol=Nj),Pf-(alpha*xmat))-xmat
  return(as.vector(y))
}
```

To run Filzbach using this likelihood function and the above data we use the `FilzbachR` package. There is a somewhat out of date version avaliable here: [http://research.microsoft.com/en-us/downloads/1cfe685f-d31a-4013-b8bb-a376d0a03480/](http://research.microsoft.com/en-us/downloads/1cfe685f-d31a-4013-b8bb-a376d0a03480/) which does not work on R 3.0.0 or above. I received an updated version from the principle author of the package, [Vassily Lyutsarev](mailto:vassilyl@microsoft.com). I have been assured by the authors that a new version will be made available online soon, along with, hopefully, a version that works on platforms besides Windows.

```{r run filzbach, cache=TRUE}
library(filzbach)
## make wrapper function to translate filzbach parameters into likelihood function parameters
## here we are going to use two dimensions for the latent niche traits (V1 and V2; Phi1 and Phi2) 
ll.wrap <- function(PhyEff, Phi1, Phi2, E1, E2, Sig, Sigi, Sigj, Ni, Nj, Nk, comm, phylo) {
  params<-list(PhyEff=PhyEff, Phi=list(Phi1=Phi1, Phi2=Phi2), E=list(E1=E1, E2=E2), Sig=Sig, Sigi=Sigi,
               Sigj=Sigj, Ni=Ni, Nj=Nj, Nk=Nk)
  LL.PLNT(params, comm, phylo)
}

## setup fixed parameters and data

comm <- fishPA[,-1]
Nk <- 2
Ni <- ncol(comm)
Nj <- nrow(comm)

## and here is the money function. Note: In R, skip the "Parameter Name" in create_parameter from C code:
## void parameter_create(char name[], double lb, double ub, double val, int type, int
## fixed, int dsply);
## void parameter_create_vector(char name[], double lb (lower bound), double ub (upper bound), 
## double val (initial value), int type (0 = -inf:inf; 1 = 0:inf), int fixed (0 = free param; 1 = fixed; 
## -1 = free but use val), int dsply (1 = output value), int number (number of params to make));
mod <- filzbach(100, 100, 
                function(PhyEff, Phi1, Phi2, E1, E2, Sig, Sigi, Sigj, comm, phylo) {
                        params<-list(PhyEff=PhyEff, Phi=list(Phi1=Phi1, Phi2=Phi2), E=list(E1=E1, E2=E2), 
                                     Sig=Sig, Sigi=Sigi, Sigj=Sigj, Ni=Ni, Nj=Nj, Nk=Nk)
                        LL.PLNT(params, comm, phylo) },
                94,
                list(
                    PhyEff = c(0.00010, 1, 1, 1, 0, 1),
                    Phi1 = c(-100, 100, 0.01, 0, 0, 1, Ni)
                    Phi2 = c(-100, 100, 0.01, 0, 0, 1, Ni)
                    E1 = c(-100, 100, 0.01, 0, 0, 1, Nj)
                    E2 = c(-100, 100, 0.01, 0, 0, 1, Nj)
                    Sig = c(0.00010, 50.0, 0.5, 1, 0, 1)
                    Sigi = c(0.00010, 50.0, 0.5, 1, 0, 1)
                    Sigj = c(0.00010, 50.0, 0.5, 1, 0, 1)
                ))


```
